Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics -- Appendix

### 2019-08-15 Appendix/appendix.Rmd
Removing color, splitting off appendix.

---
title: Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics

@@

keywords:
- scagnostics
- statistical graphics
- data visualisation
- exploratory data analysis
- data science 
- guided tour

abstract: |
  Projection pursuit is used to find interesting low-dimensional projections of high-dimensional data by optimizing an index over all possible projections. Most indexes have been developed to detect departure from known distributions, such as normality, or to find separations between known groups. Here, we are interested in finding projections revealing potentially complex bivariate patterns, using new indexes constructed from scagnostics and a maximum information coefficient, with a purpose to detect unusual relationships between model parameters describing physics phenomena. The performance of these indexes is examined with respect to ideal behaviour, using simulated data, and then applied to problems from gravitational wave astronomy. The implementation builds upon the projection pursuit tools available in the R package, tourr, with indexes constructed from code in the R packages, scagnostics, minerva and mbgraphic.

bibliography: bibliography.bib
output: rticles::springer_article
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{xcolor}
  - \usepackage{bm}
---

@@

# Introduction

<!-- 
Main point of paper:

- Some metrics for exploring high-dimensional data by finding interesting pairs of variables, can be used to find interesting projections, but some have problems. 
- What are the ways to evaluate.
- What adjustments can be done.
- How it can be applied

Organisation of this section:

- Motivation first?
- PP
- Scagnostics
- section needs shortening
-->

The term "projection pursuit" (PP) was coined by @FT74 to describe a procedure for searching high (say $p-$)dimensional data for "interesting" low-dimensional projections ($d=1$ or $2$ usually). The procedure, originally suggested by @kr69, involves defining a criterion function, or index, that measures the "interestingness" of each $d$-dimensional projection of $p$-dimensional data. This criterion function is optimized over the space of all $d$-dimensional projections of $p$-space, searching for both global and local maxima. It is hoped that the resulting solutions reveal low-dimensional structure in the data not found by methods such as principal component analysis. Projection pursuit is primarily used for visualization, with the projected data always reported as plots.

A large number of projection pursuit indexes have been developed, primarily based on departure from normality, which includes clusters, outliers and skewness, and also for finding separations between known groups (e.g. @f87, @hall89 @CBC92, @naito1997, @lckl2005, @AHC02, @CEM:CEM2568, @JS87, @5508437, @PAN2000153, @FER13, @LOPERFIDO201842).  Less work has been done on indexes to find nonlinear dependence between variables, focused on $d=2$, which motivates this research.

<!-- check Perisic and Posse (2012) and recent paper by Austrian guys -->

The driving application is from physics, to aid the interpretation of model fits on experimental results. A physical model can be considered to be a set of $p$ free parameters, that cannot be measured directly and are determined by fitting a set of $q~ (p<q)$ experimental observations, for which predictions can be made once the $p$ parameters are estimated. (Note here, that while we may have analytic expressions for the predictions, this is not always the case and we often have to rely on numerical computation.) Different sets of model parameters ($n$) found to be  compatible with the experimental results within a selected level of confidence yield the data to be examined using projection pursuit. A single prediction can be a complicated function of all of the free parameters, and typically $q \in [100,1000]$ and $p \sim 10$. Current practice is to examine pairs of parameters, or combinations produced by intuition or prior knowledge. This begs the question, whether important nonlinear associations are missed because because they are hidden in linear combinations of more than two variables.

\textcolor{red}{PP can be combined with other dimension reduction methods when $p$ is very high. For example, it can be beneficial to first do principal component analysis prior to PP, especially to remove linear dependencies before searching for other types of association. This is the approach used in @Cook:2018mvr, which explores a 56-dimensional parameter space, by first reducing the number of dimensions to the first six principal components, before applying projection pursuit. PCA was appropriate for this problem because reducing to principal component space removed the linear dependencies while preserving the nonlinear relationships that were interesting to discover. Some projection pursuit indexes do incorporate penalty terms to automate removing noise dimensions. It can also be important to have an efficient PP optimizer, particularly when working with high dimensions, because the search space increases exponentially with dimension.}

To find appropriate projections pursuit indexes for detecting nonlinear dependencies, the literature on variable selection was a starting point. With high-dimensional data, even plotting all pairs of variables can lead to too many plots, which is what "scagnostics" (@scag, @WW08) were developed to address by providing metrics from which to select the most interesting variable pairs. There are eight scagnostics, of which three ("convex", "skinny" and "stringy") are used here. The question is whether these can be adapted into projection pursuit indexes, to search for unusual features in two-dimensional projections of high-dimensional data. Recent PhD research by @Grimm2016 explored the behavior of scagnostics for selecting variables, and proposed two more that have nicer properties, based on smoothing splines and distance correlation. In addition, two more indexes for measuring dependence have been proposed in the machine learning literature, based on information criteria, maximal and total information coefficient (MIC and TIC) [@Reshef1518], with computationally more efficient versions (MIC_e, TIC_e) [@JMLRv1715308]. These are related to original 1D projection pursuit indexes based on entropy (e.g. @huber85, @JS87). This provides seven current indexes for measuring dependence between two variables, and each is available in an R [@rref] package: scagnostics [@HWscagR], mbgraphic [@mbgraphic] and minerva [@minerva].

PP index behavior can be understood and investigated more substantially when combined with a tour. A tour (@As85, @BCAH05) displays a smooth sequence of low dimensional projections from high dimensions to explore multivariate data for structure such as clusters, outliers, and nonlinear relationships. @CBCH94 provided an approach combining the tour algorithm with PP, to interactively both search for interesting projections, and examine the behavior of the indexes. The projection pursuit guided tour is available in the R package, tourr [@tourr], and provides optimization routines, and visualization.

\textcolor{red}{This paper is structured as follows.} Section \ref{sec:construct} discusses index construction, and how they can be used in the guided tour. Section \ref{sec:investigate} investigates the behavior of the indexes, explored primarily using tour methods. The new guided tour with these indexes is applied to two examples from gravitational wave astronomy (Section \ref{sec:phys}). \textcolor{red}{The latter two parts are connected in that the application of the new indexes to these problems is the main motivation for the paper, and the simulation study, in the first part, was conducted to better understand the behavior of the indexes in general. The techniques in Section \ref{sec:investigate} define procedures that will be generally useful for researchers developing new projection pursuit indexes to visually assess their behavior. Visual methods to diagnose the index behavior is important because PP is primarily used for visualistion.  The paper finishes with a discussion about the limitations of this work, and the potential future directions.}

# Projection pursuit index construction and optimization
\label{sec:construct}

A projection pursuit index (PPI) is a scalar function $f$ defined on an $d$-dimensional data set, computed by taking a $d$-dimensional projection of an $n\times p$ data matrix. Typically the definition is such that larger values of $f$ indicate a more interesting distribution of observations, and therefore maximizing $f$ over all possible $d$ dimensional projections of a data set with $p>d$ variables will find the most interesting projections. This section describes the seven indexes that are to be used to explore bivariate association. Some data pre-processing, including standardization, is advisable, prior to optimizing the PPI.

## Scaling and standardization

Making a plot always involves some choice of scaling. When a scatterplot is made, effectively, albeit under the hood, the data is scaled into a range of $[a, b]$ (often $a=0, b=1$) on both axes to print it on a page or display in a graphics device window. The range deliminates page space within which to draw. The upshot is that the original data scale is standardized to the range and aspect ratio on the display space. It may be that the original range of one variable is $[1, 1000]$ and the other is $[1, 1.6]$ but the display linearly warps this to $[0, 1]$ and $[0, 1]$, say, giving both variables equal visual weight.

With high dimensional data, and particularly projections, it is also necessary to re-scale the original range, and it is important to pay attention to what is conventional, or possible, and the effects. The PPIs also may require specific scaling for them to be effectively computed. Both of these are addressed here. The common pre-processing include:

- Standardizing each variable, to have mean 0 and variance 1, so that individual variable scales do not affect the result. Different variable scales are examinable without resorting to projection pursuit, so can be handled prior to searching through high dimensions.
- Sphering the high-dimensional data is often done to remove linear dependence. This is typically done using principal component analysis, and using the principal components as the variables passed to PP. If linear dependence is the only structure PP is not needed, and thus this is removed before PP so that the PPIs are not distracted by simple structure.
- Transform single variables to reduce skewness. It is marginal structure, visible in a single variable, which doesn't need a multivariate technique to reveal. Skewed distributions will inadvertently affect the PPIs, distracting the search for dependence. 
- Remove outliers, which may be an iterative process, to discover, identify and delete. Extreme values will likely affect PPI performance. Outliers can be examined on a case by case basis later.
- Possibly remove noise dimensions, which is also likely to be an iterative process. Directions where the distribution is purely noise make optimization of a PPI more difficult. If a variable is suspected to have little structure and relationship with other variables, conducting PP on the subset of variables without them may improve the efficiency of the search.
- Centering and scaling of the projected data, can be helpful visually. If the data has a small amount of non-normal distribution in some directions, the projected data can appear to wander around the plot window during a tour. It doesn't matter what the center of the projected data is, so centering removes a wandering scatterplot. Less commonly, it may be useful to scale the projected data to standard values, which would be done to remove any linear dependence remaining in the data.

\begin{table}[htp]
\caption{Summary of notation}
\centering
\begin{tabular}{|l|p{8.2cm}|} \hline
$p$ & be the number of variables in the data. For physics models the observable space is higher dimension, say $q$, and the data examined is the fitted model space, typically less than 10. \\
$d (=2) $ & projection dimension. For studying physics models, typically $d=2$ and this is the focus for the index definition. \\
$n$ & Number of observations, which for the physics models is the number of fitted models being examined and compared. \\
&\\
$\bm{X}_{n\times p} = (\bm{X}_1, ..., \bm{X}_p)$ & $n\times p$-dimensional data matrix, where variables $\bm{X}_j$ may be scaled or standardized, and $\bm{X}$ may be sphered\\
$\bm{Y}_{n\times 2} = (\bm{Y}_1, ..., \bm{Y}_2)$ & projected data matrix. where $Y_j = (\alpha_1 X_1, ..., \alpha_p X_p)$ \\
$\bm{F}_{p\times 2} = (\bm{\alpha}_1, ..., \bm{\alpha}_2)$ & orthonomal projection matrix\\ 
&\\
$H$ & Convex hull \\
$A$ & Alpha hull \\\hline
\end{tabular}
\label{notation}
\end{table}

## New projection pursuit index functions
\label{sec:indexDef}

Table \ref{notation} summarizes the notation used for this section. Here we give an overview of the functions that are converted into projection pursuit indexes. Full details of the functions can be found in the original sources.

* **scagnostics**: The first step to computing the scagnostics is that the bivariate data is binned, and scaled between [0,1] for calculations. The convex [@Eddy:1977] and alpha hulls [@1056714], and the minimal spanning tree (MST) [@2033241], are computed. 
    + *convex*: The ratio of the area of alpha to convex hull, $I_{\bm{F}, convex}= \frac{area(A(Y))}{area(H(Y))}$. This is the only measure where interesting projections will take low values, with a maximum of 1 if both areas are the same. Thus $1-c_{convex}$ is used.
    + *skinny*: The ratio of the perimeter to the area of the alpha hull, $I_{\bm{F}, skinny} = 1 -  \frac{\sqrt{4\pi area(A)}}{perimeter(A)}$, where the normalization is chosen such that $I_{\bm{F}, skinny} = 0$ for a full circle. Values close to 1 indicate a skinny polygon.
    + *stringy*: Based on the MST, $I_{\bm{F}, stringy} = \frac{diameter(MST)}{length(MST)}$ where the diameter is the longest connected path, and the length is the total length (sum of all edges). If the MST contains no branches $I_{\bm{F}, skinny} = 1$.

* **association**: The index functions are available in @mbgraphic, and are defined to range in [0,1]. Both functions in the mbgraphic package can bin the data before computing the index, for computational performance.
    + *dcor2D*: This function is based on distance correlation [@szekely2007], which is designed to find both linear and nonlinear dependencies between variables. It involves computing the distances between pairs of observations, conducting an analysis of variance type breakdown of the distances relative to each variable, and the result is then passed to the usual co variance and hence correlation formula. The function wdcor, in the R package, extracat [@extracat], computes the statistic, and the mbgraphic package utilises this function. 
    + *splines2D*: Measures nonlinear dependence by fitting a spline model [@wahba1990spline] of $\bm{Y}_2$ on $\bm{Y}_1$ and also $\bm{Y}_1$ on $\bm{Y}_2$, using the \texttt{gam} function in the R package, mgcv [@w16]. The index compares the variance of the residuals:
    \begin{equation}
    I_{\bm{F}, splines2d} = max(1- \frac{Var(res_{\bm{Y}_1\sim \bm{Y}_2})}{Var(\bm{Y}_1)}, 1-\frac{Var(res_{\bm{Y}_2\sim {\bm{Y}_1}})}{Var(\bm{Y}_2)}),
    \end{equation}
    which takes large values if functional dependence is strong.

* **information**:  The index functions [@Reshef1518] nonparametricly measure nonlinear association by computing the mutual information, 
\begin{equation}
\bm{I} = \sum_{by_1}\sum_{by_2} p(by_1, by_2) log(p(by_1, by_2)/(p(by_1)p(by_2))),
\end{equation}
where $by_1, by_2$ are binned values of the projected data, and $p(by_1, by_2)$is the relative bin count in each cell, and $p(by_1), p(by_2)$ are the row and column relative counts, on a range of bin resolutions of the data. It is strictly a 2D measure. For a fixed binning, e.g. $2\times 2$ or $10\times 4$, the optimal binning is found by maximizing $I$. The values of I range between $[0,1]$ because they are normalized across bins by dividing by $log(min(\# bins_{y_1}, \# bins_{y_2}))$.  
    - *Maximum Information Coefficient (MIC)*: uses the maximum normalized $I$ across all bin resolutions. 
    - *Total Information Coefficient (TIC)*: sums the normalized $I$ for all bin resolutions computed. This creates a problem of scaling - there is no upper limit, although it is related to number of bins, and number of bin resolutions used. In the work below we have made empirical estimates of the maximum and scaled the TIC index using this to get it in the range $[0,1]$. This index should be more stable than MIC.

\textcolor{red}{A comparison between these indexes for the **purpose of variable selection**, but not projection pursuit, was discussed in @Grimm2016. It is only available in German, we we summarise the main findings here. The scagnostics measures are flexible, and calculating the full set of measures provides useful guidance in variable selection. However, they are found to be highly sensitive to outlying points and sample size (as a consequence of the binning). Both splines2D and dcor2D are found to be robust in this respect, but splines2D is limited to functional dependence, while dcor2D is found to take large values only in scenarios with large linear correlation. The mutual information based index functions (MIC< TIC) are found to be flexible, but are sensitive to the sample size and often take relatively large values even when no association is present. A brief comparison of MIC and dcor2D was also provided in @2014arXiv1401.7645S.}

\textcolor{red}{In addition to the seven indexes described above, we will also include the holes index available in the `tourr` package, see [@CBC93,@Cook:2007:IDG:1557227]. This serves as a benchmark, demonstrating some desired behavior. The index takes maximum values for a central hole in the distribution of the projected data.}

## Optimization
Given a PPI we are confronted with the task of finding the maximum over all possible $d$ dimensional projections. One challenge is to avoid getting trapped in local maxima that are only a result of sampling fluctuations or a consequence of a noisy index function. @posse95b discusses the optimization, in particular that for most index functions and optimizers results are too local, largely dependent on starting point. @f87 suggested a two-step procedure: the first step is using a large step size to find the approximate global maximum while stepping over pseudomaxima. A second step is then starting from the projection corresponding to the approximate maximum and employing a gradient directed optimization for the identification of the maximum. For exploring high-dimensional data, it can be interesting to observe local maxima as well as a global maximum, and thus a hybrid algorithm that still allows lingering but not being trapped by local maxima is ideal. In addition, being able to visually monitor the optimization and see the optimal projection in the context of neighboring projections is useful. This is provided by combining projection pursuit with the grand tour [@CBCH94]. The properties of a suitable optimization algorithm include monotonicity of the index value, a variable step-size to avoid overshooting and to increase the chance of reaching the nearest maximum, and a stopping criterion allowing to move out of a local maximum and into a new search region [@tourr]. A possible algorithm is inspired by simulated annealing and has been described in @lckl2005, this has been implemented in the \texttt{search\_better} and \texttt{search\_better\_random} search functions in the tourr package. The tourr package also provides the \texttt{search\_geodesic} function, which first selects a promising direction by comparing index values between a selected number of small random steps, and then optimizes the function over the line along the geodesic in that direction considering projections up to $\pi/4$ away from the current view.

@@

# Investigation of indexes
\label{sec:investigate}

A useful projection pursuit index needs to have several properties. This has been discussed in several seminal papers, e.g. @diaconis84, @huber85, @JS87, @posse95b, @hall89. The PPI should be  minimized by the normal distribution, because this is not interesting from a data exploration perspective. If all projections are normally distributed, good modelling tools already exist. A PPI should be approximately affine invariant, regardless how the projection is rotated the index value should be the same, and the scale of each variable shouldn't affect the index value. Interestingly the original index proposed by @FT74 was not rotationally invariant. A consistent index means that small perturbations to the sample do not dramatically change the index value. This is particularly important to making optimization feasible, small angles between projections correspond to small perturbations of the sample, and thus should be small changes to index value. @posse95b suggests that indexes should be resistant to features in a tail of the distribution, but this is debatable because one departure from normality that is interesting to detect are anomalous observations. Some PPI are designed precisely for these reasons. Lastly, because we need to compute the PPI over many projections, it needs to be fast to compute. These form the basis of the criteria upon which the scagnostic indexes, and the several alternative indexes are examined, as explained below.

- **smoothness**: This is the consistency property mentioned above. The index function values are examined over interpolated tour paths, that is, the value is plotted against time, where time indexes the sequence of projections produced by the tour path. The signature of a good PPI is that the plotted function is relatively smooth. The interpolation path corresponds to small angle changes between projections, so the value should be very similar.

- **squintability**: @barnett1981interpreting introduced the idea of squint angle to indicate resolution of structure in the data. Fine structure like the parallel planes in the infamous RANDU data [@Marsaglia68] has a small squint angle because you have to be very close to the optimal projection plane to be able to see the structure. Structures with small squint angle are difficult to find, because the optimization algorithm needs to get very close to begin hill-climbing to the optimum. The analyst doesn't have control over the data structure, but does have control over the PPI. Squintability is about the shape of the PPI over all projections. It should have smooth low values for noise projections and a clearly larger value, ideally with a big squint angle, for structured projections. The optimizer should be able to clearly see the optimal projections as different from noise. To examine squintability, the PPI values are examined on interpolated tour paths between a noise projection and a distant structured projection.

- **flexibility**: An analyst can have a toolbox of indices that may cover the range of fine and broad structure, which underlies the scagnostics suite. Early indexes, based on density estimation could be programmed to detect fine or large structure by varying the binwidth. This is examined by using a range of structure in the simulated data examples.

- **rotation invariance**: The orientation of structure within a projection plane should not change the index value. This is especially important when using the projection pursuit guided tour, because the tour path is defined between planes, along a geodesic path, not bases within planes. If a particular orientation is more optimal, this will get lost as the projection shown pays no attention to orientation. @BCAH05 describes alternative interpolation paths based on Givens and Householder rotations which progress from basis to basis. It may be possible to ignore rotation invariance with these interpolations but there isn't a current implementation, primarily because the within-plane spin that is generated is distracting from a visualization perspective. Rotation invariance is checked for the proposed PPIs by rotating the structured projection, within the plane.

- **speed**: Being fast to compute allows the index to be used in real-time in a guided tour, where the optimization can be watched. When the computations are shifted off-line, to watch in replay, computation times matter less. This is checked  by comparing times for benchmark scenarios with varying sample size.

## Simulation study setup

### Data construction
\label{sec:dataOv}

Three families of data simulations are used for examining the behavior of the index functions. Each generates  structure in two variables, with the remaining variables containing various types of noise. This is a very simple construction, because there is no need for projection pursuit to find the structure, one could simply use the PPIs on pairs of variables. However, it serves the purpose to also evaluate the PPIs. The three data families are explained below. In each set, $n$ is used for the number of points, $p$ is the number of dimensions, and $d=2$ is the projection dimension. 
\textcolor{red}{The three structures were selected to cover both functional and non-functional dependence, different types of nuisance distributions and different structure size and squintability properties.}

- **pipe**: nuisance directions are generated by sampling independently from a uniform distribution between $[-1,1]$, and the circle is generated by sampling randomly on a 2D circle, and adding a small radial noise. The circle should be easy to see by some indices because it is large structure, but the nonlinearity creates a complication. 
- **sine**:  nuisance directions are generated by sampling independently from a standard normal distribution, and the sine curve is generated by $x_p = \sin(x_{p-1}) + \mathrm{jittering}$. The sine is a medium nonlinear structure, which should be visible to multiple indices.
- **spiral**:  nuisance directions are generated by sampling independently from a normal distribution, and the structure directions are sampled from an Archimedean spiral, i.e. $r = a + b \theta$, with $a=b=0.1$ and we sample angles $\theta$ from a normal distribution with mean 0 and variance $2\pi$, giving a spiral with higher densities at lower radii. The absolute value of $\theta$ fixes the direction of the spiral shape. This is fine structure which is only visible close to the optimal projection.

@@

```{r data, fig.width=6, fig.height=7, fig.cap="Scatterplots of pairs of variables from samples of each family, showing the nuisance variables and structured variables.", out.width = "0.8\\textwidth"}
dsn <- spiral1000 %>% select(V1, V2) %>%
  rename(X1=V1, X2=V2) %>%
  mutate(family="spiral", type="nuisance")
dss <- spiral1000 %>% select(V5, V6) %>%
  rename(X1=V5, X2=V6) %>%
  mutate(family="spiral", type="structured")
dpn <- pipe1000 %>% select(V1, V2) %>%
  rename(X1=V1, X2=V2) %>%
  mutate(family="pipe", type="nuisance")
dps <- pipe1000 %>% select(V5, V6) %>%
  rename(X1=V5, X2=V6) %>%
  mutate(family="pipe", type="structured")
din <- sin1000 %>% select(V1, V2) %>%
  rename(X1=V1, X2=V2) %>%
  mutate(family="sine", type="nuisance")
dis <- sin1000 %>% select(V5, V6) %>%
  rename(X1=V5, X2=V6) %>%
  mutate(family="sine", type="structured")
d <- bind_rows(dsn, dss, dpn, dps, din, dis)
ggplot(d, aes(x=X1, y=X2)) + geom_point(alpha=0.3) + 
  facet_grid(type~family) + theme(aspect.ratio=1) +
  xlab("") + ylab("")
```

```{r indexTable, echo=FALSE}
tableTemp <- function(){
  tibble(
    index=character(),
    data=character(),
    size=numeric(),
    type=character(),
    value5=numeric(),
    value95=numeric()
  )
}

getSample <- function(dataT, size){
  if (dataT == "Pipe"){
    ret <- pipeData(p = size, n = 6) %>% scale() %>% as_tibble()
  }
  if (dataT == "Spiral"){
    ret <- spiralData(p = size, n = 6) %>% scale() %>% as_tibble()
  }
  if (dataT == "Sine"){
    ret <- sinData(p = size, n = 6) %>% scale() %>% as_tibble()
  }
  return(ret)
}

getIdx <- function(dataT, size, type){
  convex <- numeric(length = 100)
  skinny <- numeric(length = 100)
  stringy <- numeric(length = 100)
  micidx <- numeric(length = 100)
  ticidx <- numeric(length = 100)
  dcor2D <- numeric(length = 100)
  splines2D <- numeric(length = 100)
  holesI <- numeric(length = 100)
  for(i in 1:100){
    cSample <- getSample(dataT, size)
    if(type == "noise"){
      x = as.numeric(cSample$V1)
      y = as.numeric(cSample$V2)
    } else{
      x = as.numeric(cSample$V5)
      y = as.numeric(cSample$V6)
    }
    scag <- scagnostics(x=x, y=y)$s
    midx <- mine(x, y, est = "mic_e")
    skinny[i] <- scag["Skinny"]
    stringy[i] <- scag["Stringy"]
    convex[i] <- scag["Convex"]
    micidx[i] <- midx$MIC
    if (size == 100) {ticidx[i] <- midx$TIC / 16}
    if (size == 1000) {ticidx[i] <- midx$TIC / 148}
    dcor2D[i] <- dcor2d(x=x, y=y)
    splines2D[i] <- splines2d(x,y)
    holesI[i] <- holes(matrix(c(x,y), ncol = 2))

  }
  #sortedList <- lapply(list(skinny, stringy, convex, micidx, ticidx, dcor2D, splines2D, holesI), sort)
  skinny <- sort(skinny)
  stringy <- sort(stringy)
  convex <- sort(convex)
  micidx <- sort(micidx)
  ticidx <- sort(ticidx)
  dcor2D <- sort(dcor2D)
  splines2D <- sort(splines2D)
  holesI <- sort(holesI)
  res <- tableTemp() %>%
    add_row(index="skinny", data=dataT, size=size,
            type=type,
            value5=skinny[5], value95=skinny[95]) %>%
    add_row(index="stringy", data=dataT, size=size,
            type=type,
            value5=stringy[5], value95=stringy[95]) %>%
    add_row(index="convex", data=dataT, size=size,
            type=type,
            value5=convex[5], value95=convex[95]) %>%
    add_row(index="MICe", data=dataT, size=size,
            type=type,
            value5=micidx[5], value95=micidx[95]) %>%
    add_row(index="TICe", data=dataT, size=size,
            type=type,
            value5=ticidx[5], value95=ticidx[95]) %>%
    add_row(index="dcor2D", data=dataT, size=size,
            type=type,
            value5=dcor2D[5], value95=dcor2D[95]) %>%
    add_row(index="splines2D", data=dataT, size=size,
            type=type,
            value5=splines2D[5], value95=splines2D[95]) %>%
    add_row(index="holes", data=dataT, size=size,
            type=type,
            value5=holesI[5], value95=holesI[95])
}

if(!file.exists("cache/indexTable.rda")){
  set.seed(505)
  res <- tableTemp()
  for(ds in c("Pipe", "Sine", "Spiral")){
    for (size in c(100, 1000)){
      for (type in c("structure", "noise")){
        res <- bind_rows(res,getIdx(ds, size, type))
      }
    }
  }
  res <- res %>% arrange(data, size, index, type) %>% select(data, size, index, type, value5, value95)
  save(res, file = "cache/indexTable.rda")
} else {
  load("cache/indexTable.rda")
}
#rescale holes index first
res <- res %>%
  rowwise() %>%
  mutate(value5 = if_else(index=="holes", rescaleHoles(value5), value5)) %>%
  rowwise() %>%
  mutate(value95 = if_else(index=="holes", rescaleHoles(value95), value95))
resp <- res %>% filter(size == 1000, data == "Pipe") %>%
  select(index, type, value5, value95) %>%
  rename(lower=value5, upper=value95)
resi <- res %>% filter(size == 1000, data == "Sine") %>%
  select(index, type, value5, value95) %>%
  rename(lower=value5, upper=value95)
ress <- res %>% filter(size == 1000, data == "Spiral") %>%
  select(index, type, value5, value95) %>%
  rename(lower=value5, upper=value95)
res_all <- bind_cols(resp, resi, ress) %>%
  select(-index1, -type1, -index2, -type2) %>%
  mutate(indexF = factor(index, c("holes", "convex", "skinny", "stringy", "dcor2D", "splines2D", "MICe", "TICe"))) %>%
  arrange(indexF) %>%
  select(-indexF, -index) %>%
  add_column(index = c("holes", "", "convex", "", "skinny", "", "stringy", "", "dcor2D", "", "splines2D", "", "MICe", "", "TICe", "")) %>%
  select(index, everything())
# Good resource for doing tables
# https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_pdf.pdf
kable(res_all, format = "latex",  caption = "Comparison of index values between noise projections and structured projections for sample size 1000, using 5th and 95th percentiles from 100 simulated sets. Most indices have much larger values for the structured projections, except for convex.", 
      col.names=c("Index","", "lower", "upper", "lower", "upper", "lower", "upper"),
      digits=2, booktabs = T) %>% 
      add_header_above(c(" ", " ", "pipe" = 2, "sine" = 2, "spiral"=2)) %>%
  group_rows("", 1, 2, latex_gap_space = "-0.2cm") %>%
  group_rows("", 3, 4, latex_gap_space = "-0.2cm") %>%
  group_rows("", 5, 6, latex_gap_space = "-0.2cm") %>%
  group_rows("", 7, 8, latex_gap_space = "-0.2cm") %>%
  group_rows("", 9, 10, latex_gap_space = "-0.2cm") %>%
  group_rows("", 11, 12, latex_gap_space = "-0.2cm") %>%
  group_rows("", 13, 14, latex_gap_space = "-0.2cm") %>%
  group_rows("", 15, 16, latex_gap_space = "-0.2cm") %>% 
  #row_spec(1:16, font_size = 10) %>%
  kable_styling(position = "center") #, font_size = 7) 
```

\textcolor{red}{For simplicity, in the investigations of the index behavior, we fix $p=6$, which corresponds two independent 2D planes containing nuisance distributions, and one 2D plane containing the structured distribution. }
The structured projection is in variables 5 and 6 ($x_5, x_6$). Two samples sizes are used: $n=(100, 1000)$. All variables are standardized to have mean 0 and standard deviation 1. Figure \ref{fig:data} shows samples from each of the families, of the nuisance and structured pairs of variables. Table \ref{tab:indexTable} compares the PPIs for structured projections against those for nuisance variables, based on 100 simulated data sets of each type, using sample size 1000. The lower and upper show the 5th and 95th percentile of values. The holes index is sensitive only to the pipe distribution. All other indexes, except convex show distinctly higher values for the structured projections. The convex index shows the inverse scale to other indices, thus (1-convex) will be used in the assessment of performance of PPIs. The scale for the holes index in its original implementation is smaller than the others ranging from about 0.7 through 1, so it is re-scaled in the performance assessment so that all indices can be plotted on a common scale of 0-1 (details in Appendix). Similarly, the TIC index is re-scaled depending on sample size.

### Property assessment {#sec:propAss}

The procedures for assessing the PPI properties of smoothness, squintability, flexibility, rotation invariance, and speed examined for samples from the family of data sets are:

1. Compute the PPI values on the tour path along an interpolation between pairs of nuisance variables, $x_1 - x_2$ to $x_3 - x_4$. The result is ideally a smooth change in low values. This checks the smoothness property. 
2. Change to a tour path between a pair of nuisance variables $x_1 - x_2$ and the structured pair of variables $x_5 - x_6$ via the intermediate projection onto $x_1 - x_5$, and compute the PPI along this. This examines the squintability, and smoothness. If the function is smooth and slowly increases towards the structured projection, then the structure is visible from a distance. 
3. Use the guided tour to examine the ease of optimization. This depends on having a relatively smooth function, with structure visible from a distance. One index is optimized to show how effectively the maximum is attained, and the values for other PPIs is examined along the same path, to examine the similarity between PPIs. 
4. Rotation invariance is checked by computing PPIs on rotations of the structured projection. 
5. Computational speed for the selected indexes is examined on a range of sample sizes.

```{r dataPlotsV1V2, fig.width=6, fig.height=7, fig.cap="Projection of all data sets considered onto the first two parameters $x_1$ and $x_2$, i.e. showing a typical uninformed view for each dataset.", eval=FALSE}
pL <- list()
i <- 1
pLabels <- c("Spiral 100", "Spiral 1000", "Pipe 100", "Pipe 1000", "Sine 100", "Sine 1000")
for(ds in list(spiral100, spiral1000, pipe100, pipe1000, sin100, sin1000)){
  pC <- ggplot(ds, aes(V1, V2)) +
    geom_point() +
    ggtitle(pLabels[i]) + theme(aspect.ratio=1)
  pL[[i]] <- pC
  i <- i+1
}
grid.arrange(pL[[1]], pL[[2]], pL[[3]], pL[[4]], pL[[5]], pL[[6]],
             ncol=2)
```

```{r dataPlotsV5V6, fig.width=6, fig.height=7, fig.cap="Projection of all data sets considered onto the last two parameters $x_5$ and $x_6$, i.e. showing the special view built into the datasets.", eval=FALSE}
pL <- list()
i <- 1
pLabels <- c("Spiral 100", "Spiral 1000", "Pipe 100", "Pipe 1000", "Sine 100", "Sine 1000")
for(ds in list(spiral100, spiral1000, pipe100, pipe1000, sin100, sin1000)){
  pC <- ggplot(ds, aes(V5, V6)) +
    geom_point() +
    ggtitle(pLabels[i]) + theme(aspect.ratio=1)
  pL[[i]] <- pC
  i <- i+1
}
grid.arrange(pL[[1]], pL[[2]], pL[[3]], pL[[4]], pL[[5]], pL[[6]],
             ncol=2)
```

```{r dataPlotsDensity, fig.width=6, fig.height=7, fig.cap="One dimensional density distribution for all parameters and all data sets considered. Apart from statistical fluctuations we see that for the Pipe datasets the parameters $x_5$ and $x_6$ are pushed towards more extreme values by vetoing points away from the circle outline, and for the Sine (and Spiral) dataset the parameter $x_6$ (and $x_5$) follows a very distinct distribution.", eval=FALSE}
pL <- list()
i <- 1
pLabels <- c("Spiral 100", "Spiral 1000", "Pipe 100", "Pipe 1000", "Sine 100", "Sine 1000")
for(ds in list(spiral100, spiral1000, pipe100, pipe1000, sin100, sin1000)){
  distData <- gather(ds, variable, value)
  pC <- ggplot(distData, aes(value, color=variable)) +
    geom_density() +
    ggtitle(pLabels[i])
  pL[[i]] <- pC
  i <- i+1
}
grid.arrange(pL[[1]], pL[[2]], pL[[3]], pL[[4]], pL[[5]], pL[[6]],
             ncol=2)
```

## PPI traces over a tour sequence of interpolated nuisance projections
\label{sec:smooth}

Figure \ref{fig:plotV1V2toV3V4} shows the traces representing the index values calculated across a tour between a pair of nuisance projections.
\textcolor{red}{The tour path is generated by interpolating between the two independent nuisance planes, i.e. from the projection onto $x_1$-$x_2$ to one onto $x_3$-$x_4$.}
The range of each axis is set to be the limits of the index, as might be expected over many different data sets, 0 to 1. Each projection in the interpolation will also be noise. Two different sample sizes are show, $n=100$ as a dashed line, and $n=1000$ as a solid line. The ideal trace is a smooth function, with relatively low values, and no difference between the sample sizes. A major feature to notice is that the scagnostics produce noisy functions, which is problematic, because small changes in the projection result in big jumps in the value. This will make them difficult to optimize. On the other hand holes, dcor2d, splines2d, MIC and TIC are relatively smooth functions.

Several of the indexes are sensitive to sample size also, the same structured projection with differing numbers of points, produces different values.

>>>>

---
title: Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics -- Appendix

@@

abstract: |
  Appendix material for the paper "Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics".

bibliography: ../bibliography.bib
output: rticles::springer_article
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{xcolor}
  - \usepackage{bm}
---

@@

@@

@@
